---
title: "Dependencies Installation Guide"
vignette: >
  %\VignetteIndexEntry{Dependencies Installation Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

Many of the required packages can be installed in more than one way, and only a few are hosted on CRAN. For each package, we give our recommended installation method and link to its documentation for alternatives.

## R & RStudio Installation

Install the latest version of R & RStudio for your device

- [R](https://www.r-project.org/)
- [RStudio](https://posit.co/products/open-source/rstudio)

## Rtools Installation (Windows Only)

If you're using a Windows device, download Rtools before installing the following dependencies

- [RTools](https://cran.rstudio.com/bin/windows/Rtools/)

## Dependencies Installation

To install and use `phylodyn` & `phylodyn2`, you'll need to install the following dependencies.

### From 

These are the necessary packages from CRAN:
```{r}
#| eval: false
install.packages(c("ape", "spam", "loo", "rstan", "pak"))
```

### From GitHub

```{r}
#| eval: false
pak::pak("jrfaulkner/spmrf")
```

### External Packages 

You'll need to install `Rgraphviz` and `graph` using:
```{r}
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c("graph", "Rgraphviz"))
```
*See [Bioconductor](https://www.bioconductor.org/packages/devel/bioc/html/Rgraphviz.html) for package documentation.*

### INLA Installation

Download `INLA` using:
```{r}
#| eval: false
install.packages(
  "INLA", 
  repos = c(
    getOption("repos"), 
    INLA = "https://inla.r-inla-download.org/R/stable"
  ), 
  dep = TRUE
)
```

If the *stable* installation does not work, use the *testing* version instead 
```{r}
#| eval: false
install.packages(
  "INLA", 
  repos = c(
    getOption("repos"), 
    INLA = "https://inla.r-inla-download.org/R/testing"
  ), 
  dep = TRUE
)
```
*See [inla.org](https://www.r-inla.org/download/index.html) for package documentation.*

## Phylodyn Installation

After installing all dependencies, (adding pak) install `phylodyn` using:
```{r}
#| eval: false
pak::pak("mdkarcher/phylodyn")
```
*See <https://github.com/mdkarcher/phylodyn> for package documentation.* 

## Phylodyn2 Installation

Install using: 
```{r}
#| eval: false
pak::pak("CatalinaMedina/phylodyn2")
```
*See <https://github.com/CatalinaMedina/phylodyn2> for package documentation.*

## Adapref Installation 

Make sure all dependencies have been installed. Install adapref using:
```{r}
#| eval: false
pak::pak("lorenzocapp/adapref")
```
*See <https://github.com/lorenzocapp/adapref> for documentation.*

