---
title: "Bayesian Nonparametric Phylodynamic Reconstruction with reporting delays: COVID-19 in Washington"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Nonparametric Phylodynamic Reconstruction with reporting delays: COVID-19 in Washington}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows a case study of Washington SARS-CoV-2 data, using `phylodyn2`.
We start by loading the `phylodyn2` package.

```{r setup}
library(phylodyn2)
```

In preparation, we aligned the sequences using the software [MAFFT](https://mafft.cbrc.jp/alignment/software/),
and inferred a maximum clade credibility genealogy using the software [BEAST](https://beast.community/)[^1].
We packaged this genealogy in a `phylo` object `washington_genealogy`, and we load it now.

```{r}
data("washington_genealogy")
```


[^1]: We infer the genealogy branch lengths in units of years using a strict molecular clock, a Bayesian Skygrid effective population size prior, and an HKY substitution model. The code to replicate this analysis can be found at <https://github.com/CatalinaMedina/reporting-delays-in-phylodynamics-paper>.

We use `BNPR()` and `BNPR_PS()` and the `BNPR_PS_with_RD()` to calculate approximate effective population size marginals (without and with a sampling model). 
The `BNPR_PS_with_RD()` uses the delay aware preferential sampling model proposed in [TODO: add paper](TODO). 
This model needs a vector of reporting delays to calculate the empirical probability of a sample being reported by time zero of the analysis.

The `washington_delays` object is a data frame of sampling and reporting dates for a random subset of SARS-CoV-2 sequences in the state of Washington reported in July of 2021 to [GISAID](https://gisaid.org/). 
August 1st is the "time zero" for our analysis so these dates are taken to be representative of recent reporting delays.

```{r}
data("washington_delays")
```

Add a column for each reporting delay

```{r}
recent_delays <- as.numeric(
  washington_delays$reporting_date - washington_delays$sampling_date
)

time_zero_date <- as.Date("2021-08-01")
```

We are now ready to calculate the effective population size marginals with each of the three methods available in phylodyn.

```{r}
# Approx grid width of 3 days. See paper repo for details.
grid_length <- 163 

# Preferential sampling coefficient prior mean
pref_samp_coef_mean <- 1

# Preferential sampling coefficient prior precision
pref_samp_coef_precision <- 1 / (0.5)^2

# Time (years) between time zero of analysis and first observed sampling time. 
# This is required to ensure correct matching between sampling time and 
# reporting probability. See paper repo for details.
time_zero_offset <- 0.03011636 

set.seed(1520)
bnpr_fit <- BNPR(
  washington_genealogy, 
  lengthout = grid_length
)

set.seed(1520)
bnpr_ps_fit <- BNPR_PS(
  washington_genealogy, 
  lengthout = grid_length,
  beta1_mean = pref_samp_coef_mean, 
  beta1_prec = pref_samp_coef_precision
)

set.seed(1520)
bnpr_ps_with_rp_fit <- BNPR_PS_with_RD(
  washington_genealogy, 
  lengthout = grid_length,
  historic_reporting_delays = recent_delays / 365.25,
  beta1_mean = pref_samp_coef_mean, 
  beta1_prec = pref_samp_coef_precision,
  time_offset = time_zero_offset
)

```

Now we can plot the results to compare. 
`plot_BNPR()` could be used to plot a single trajectory.

Here we will manually plot the results for easier comparison between methods, and conversion of time to dates for x axis labels.

```{r, fig.width = 6, fig.height = 4.5, message = FALSE}
library(ggplot2)

all_output <- rbind(
  data.frame(
    t = bnpr_fit$x,
    eff_pop_median = bnpr_fit$effpop,
    eff_pop_95ub = bnpr_fit$effpop975,
    eff_pop_95lb = bnpr_fit$effpop025,
    model = "BNPR"
  ),
  data.frame(
    t = bnpr_ps_fit$x,
    eff_pop_median = bnpr_ps_fit$effpop,
    eff_pop_95ub = bnpr_ps_fit$effpop975,
    eff_pop_95lb = bnpr_ps_fit$effpop025,
    model = "BNPR PS"
  ),
  data.frame(
    t = bnpr_ps_with_rp_fit$x,
    eff_pop_median = bnpr_ps_with_rp_fit$effpop,
    eff_pop_95ub = bnpr_ps_with_rp_fit$effpop975,
    eff_pop_95lb = bnpr_ps_with_rp_fit$effpop025,
    model = "Delay Aware BNPR PS"
  )
)

# Time = 0 is first observed sampling time, but really we want it relative to our time zero of anlysis
all_output$adj_t <- all_output$t + time_zero_offset

# Change scale from time (years) to dates
all_output$Date <- as.Date(-all_output$adj_t * 365, origin = time_zero_date)

# Subset out older dates
plotting_data <- all_output[
  all_output$Date >= as.Date("2021-02-02") &
  all_output$Date <= as.Date("2021-08-01")
, ]

plotting_data |>
  ggplot(aes(x = Date, y = eff_pop_median, color = model)) +
  geom_line(linetype = "dashed") +
  geom_ribbon(
    aes(
      ymin = eff_pop_95lb, 
      ymax = eff_pop_95ub, 
      color = model, 
      fill = model
    ), 
    alpha = 0.4, 
    colour = NA
  ) +
  annotate(
    'rect',
    xmin = as.Date("2021-02-02"), 
    xmax = as.Date("2021-07-02"),
    ymin = -Inf, ymax = Inf,
    fill = '#00000010'
  ) +
  coord_cartesian(ylim = c(0, 35)) +
  labs(
    y = "Effective Population Size",
    color = "Model",
    fill = "Model"
  ) +
  scale_x_date(
    date_breaks = "1 months", 
    date_labels = "%b %d, %Y", 
    expand = c(0, 0),
    limits = c(as.Date("2021-02-02"), as.Date("2021-08-01"))
  ) +
  scale_color_manual(
    values = c("blue", "red", "darkgreen"),
    labels = c("BNPR", "BNPR PS", "Delay Aware BNPR PS")
  ) +
  scale_fill_manual(
    values = c("blue", "red", "darkgreen"),
    labels = c("BNPR", "BNPR PS", "Delay Aware BNPR PS")
  ) +
  guides(
    color = guide_legend(title.position = "top", title.hjust = 0.5),
    fill = guide_legend(title.position = "top",  title.hjust = 0.5)
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    legend.direction = "horizontal",
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    legend.key.size = unit(1, "lines")
  ) + 
  facet_wrap(
    ~  factor(
      model, 
      levels = c(
        "BNPR",
        "BNPR PS",
        "Delay Aware BNPR PS"
      )
    ), 
    nrow = 1
  )
```

The white background indicates the time period where reporting delays are likely.
The BNPR model has wider credible intervals than the other two models.
The BNPR PS and delay aware BNPR PS models have similar results in the time period without reporting delays (gray shaded background)
The delay aware BNPR PS model infers higher effective population size near time zero, August 1, than the BNPR PS model.
